Asymmetric relationships and their effects on coexistence

### 2023-08-02 paper/Paper.Rnw
typos

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study relationships, ecologists 
very commonly use metrics such as correlation. But standard measures of the strength 
and direction of association between variables do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations are also often used (e.g., various forms of 
regression), but other common methods also reveal only a small portion 
of the information content of the relationship \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently 
insights have been gained by studying how features of associations beyond standard measures 
impact ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples of ATAs in nature have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition have not been investigated. Doing so is our goal here.

@@

%Intro to storage effects and why they might be influenced by ATAs
We indicate why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects, starting by conceptually reviewing 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}. 
Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species ($E$) and the strength of competition ($C$) experienced by that species (called $EC$ covariance);
and on the assumption that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive $EC$ covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent $EC$ covariance contributes to 
differences between growth rates of currently rare versus currently common species \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on $EC$ covariation, 
and ATAs can be an aspect of $EC$ covariation. One scenario in which it is reasonable to imagine that
ATAs may be present in $EC$ covariation pertains to water limitation. Suppose two plants compete primarily for 
water, when water is scarce, but their niches become more differentiated when water is abundant.
Then the quality of the environment, $E$, as experienced by whichever plant is currently rare may only correspond closely to the 
competition, $C$, imposed on it by the currently abundant plant when water is scarce and $E$ is low.

@@

\subsection{Our decomposition as an extention of modern coexistence theory}

@@

%Setup - two versions, and basics which apply to both
For goal G2 of the Introduction, we applied our theory to the lottery model, 
comparing the magnitude of the ATA effect on coexistance, $\Delta_i^{[EC]}$, to the magnitudes 
of the other terms of 
(\ref{eq:full_decomp}) for various model parameterizations. We used two versions 
of the lottery model. The versions both used the same dynamical equations
(Introduction), but 
differed in the distributions of $(B_1,B_2)$ used. Both distributions were based on 
transformations of the distributions $(b_1,b_2)$ of Fig. \ref{fig:pedagog1}a-c;
the details of how the left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c), 
and symmetrically (Fig. \ref{fig:pedagog1}b) tail associated alternatives for $(b_1,b_2)$ were generated 
are in SI section \ref{SI-sect:noise}.

>>>>

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study relationships, ecologists 
very commonly use metrics such as correlation. But standard measures of the strength 
and direction of association between variables do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations are also often used (e.g., various forms of 
regression), but other common methods also reveal only a small portion 
of the information content of the relationship \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recent 
insights have been gained by studying how features of associations beyond standard measures 
impact ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples of ATAs in nature have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition have not been investigated. Doing so is our goal here.

@@

%Intro to storage effects and why they might be influenced by ATAs
We indicate why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects, starting by conceptually reviewing 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}. 
Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species ($E$) and the strength of competition ($C$) experienced by that species (called $EC$ covariance);
and on the assumption that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive $EC$ covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent $EC$ covariance contributes to 
differences between growth rates of currently rare versus currently common species \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on $EC$ covariation, 
and ATAs can be an aspect of $EC$ covariation. One scenario in which it is reasonable to imagine that
ATAs may be present in $EC$ covariation pertains to water limitation. Suppose two plants compete primarily for 
water, when water is scarce, but their niches become more differentiated when water is abundant.
Then the quality of the environment, $E$, as experienced by whichever plant i. s currently rare may only correspond closely to the 
competition, $C$, imposed on it by the currently abundant plant when water is scarce and $E$ is low.

@@

\subsection{Our decomposition as an extension of modern coexistence theory}

@@

%Setup - two versions, and basics which apply to both
For goal G2 of the Introduction, we applied our theory to the lottery model, 
comparing the magnitude of the ATA effect on coexistence, $\Delta_i^{[EC]}$, to the magnitudes 
of the other terms of 
(\ref{eq:full_decomp}) for various model parameterizations. We used two versions 
of the lottery model. The versions both used the same dynamical equations
(Introduction), but 
differed in the distributions of $(B_1,B_2)$ used. Both distributions were based on 
transformations of the distributions $(b_1,b_2)$ of Fig. \ref{fig:pedagog1}a-c;
the details of how the left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c), 
and symmetrically (Fig. \ref{fig:pedagog1}b) tail associated alternatives for $(b_1,b_2)$ were generated 
are in SI section \ref{SI-sect:noise}.
